---
title: "Main Gap Analysis"
author: "Tom Gibbens-Matsuyama, Emma Bea Mitchell, Karol Paya, Takeen Shamloo"
format:
  html: 
    toc: true
    toc-expand: 2
    toc-title: Gap Analysis 
    toc-location: left
editor: visual
editor_options: 
  chunk_output_type: console
---

## Introduction

This document is here to provide discrepancies in the data located on the [California Natural Resource Agency's website](https://data.cnra.ca.gov/dataset/urban-water-data-drought). The California Water Data Consortium, in partnership with the Department of Water Resources (DWR) and the State Water Resources and Control Board aim to improve urban water supplier data's accessibility and usability. In conjunction with UCSB's Bren School's Master of Environmental Data Science CalDrought Capstone Team, we aim to provide a full synthesis of the following data, providing a contextualized breakdown as well as written code for checking said breakdown. This document is broken down into five main categories representing the five main data sets used in this project. The five main data sets are as follows: `Actual Water Shortage Level`, `Five Year Water Shortage Outlook`, `Historical Production and Delivery`, `Monthly Water Shortage Outlook`, and `Source Name`.

```{r}
# load in libraries

library(tidyverse)
library(here)

# load in data

actual_water_shortage_level <- read_csv(here("data", "actual_water_shortage_level.csv"))
five_year_water_shortage_outlook <- read_csv(here("data", "five_year_water_shortage_outlook.csv"))
historical_production_delivery <- read_csv(here("data", "historical_production_delivery.csv"))
monthly_water_shortage_outlook <- read_csv(here("data", "monthly_water_shortage_outlook.csv"))
source_name <- read_csv(here("data", "source_name.csv"))
```

## Actual Water Shortage Level

### Dataset Description

## Five Year Water Shortage Outlook

### Dataset Description

## Historical Production and Delivery

### Dataset Description

```{r}
# NAs in dataset
sum(is.na(historical_production_delivery))

# negative values in quantity acre feet
negative_quantity_historical <- historical_production_delivery |> 
  filter(quantity_acre_feet < 0)

```

## Monthly Water Shortage Outlook

### Dataset Description

## Source Name

### Dataset Description
